 

### 2021-05-22 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density. This gradient is most pronounced under broad-leaf and mixed forests where light reaching forest floors, incident to canopy-tops range from ~1-2% to 4-5%, respectively [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006] compared to needle-leaf forest floors that receive ~17-34% [@baldocchiSeasonalVariationEnergy1997]. Sunflecks, offer another light source for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

@@

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ difference between understory and canopy vary based on multiple factors such as radiation, local topography, vertical air mixing and cloud cover [*Defrenne et al. 2021*].Dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies. Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009;  @martin_boundary_1999] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. (*This sentence is too complex-- hard to understand*)

@@

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a]. RH, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory. Beyond a certain tree-height, the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

@@

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994].During $gs$, evapotranspiration facilitates heat dissipation. But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].

@@

High wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, through smaller $g_a$, which increases $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]

Leaf interiors are typically $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018].

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to higher $ga$ than larger leaves, therefore maintain lower $T_{leaf}$-$T_{air}$ than large leaves that experience greater $T_{leaf}$-$T_{air}$. With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures, whereas species with larger exposed-leaves can experience greater $T_{leaf}$-$T_{air}$ and heat stress [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @bauerleSeparatingFoliarPhysiology2011; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007]

@@

The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves. $N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in sun- and, shade-leaves during sunflecks, under available soil nitrogen [@kitaoCanopyNitrogenDistribution2018].Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021], and temperate trees [@mau_temperate_2018]. *We [-nidhi, for now] hypothesize that the phenotypic-plastic nature of leaf investment in area-to-mass traits across a vertical gradient stablizes photosynthetic $T_{opt}$. Where as, across species photosynthetic $T_{opt}$ might vary more*

>>>>

**Light conditions, specifically changes in the proportion of incident light and photosynthetically active radiation (PAR), vary along the vertical gradient with leaf area index (LAI), canopy height, canopy structure, and across species and forest types [Fig. `r fig_NEON_vertical`; @koike_leaf_2001].** 
Generally, along the vertical gradient of forests globally, amount of light decreases from canopy tops to the forest floor which is mediated by foliage LAI and density. This gradient is most pronounced under dense canopies, including broad-leaf and mixed forests, than open canopy forests such as needle-leaf forests [Fig. `r fig_NEON_vertical`; @aussenacInteractionsForestStands2000a; Smith et al., 2019; Tymen et al., 2017; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019; @bartemucciOverstoryInfluencesLight2006; @baldocchiSeasonalVariationEnergy1997]. Sunflecks, offer another light source for shaded forest regions, caused by small canopy gaps and wind-induced canopy movements [@waySunflecksTreesForests2012]. In heterogeneous canopies [with high gap fractions and large variation in tree height], the distance from the outer canopy to the inner canopy is a better proxy for light environment than height (Parker 1995).

@@

### Air temperature 
**Air temperature, $T_{air}$, often shows little variation across the vertical gradient, but under certain circumstances it can be significantly buffered by forest canopies (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).**
The $T_{air}$ gradient along canopies and understories vary with factors such as radiation fluxes, local topography, vertical air mixing, cloud cover and vegetation [*Defrenne et al. 2021*]. Typically, dense canopies buffer understories from maximum $T_{air}$ relatively more than open canopies. Therefore, dense canopy understories can have cooler maximum temperature during the day, and warmer minimum temperature during the night, than open canopy understories, nearby clearings [*Defrenne et al. 2021*; @vonarxSpatiotemporalEffectsForest2012a; @defrenneGlobalBufferingTemperatures2019; @davis_microclimatic_2019; @zellweger_seasonal_2019; @ramboCanopyMicroclimateResponse2009] and neighboring tree crop agricultural plantations [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018]. (*This sentence is too complex-- hard to understand*)

@@

### Humidity and evaporative demand
**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**[@mcgregor_tree_nodate]
Along with temperature, dense-canopy forests maintain higher daily maximum relative humidity ($RH$) below-canopy than open-forests and nearby open areas, which is greater in wetter conditions, in warmer than colder months [@vonarxSpatiotemporalEffectsForest2012a]. $RH$, in combination with temperature, determines vapor pressure deficit [$VPD$]--the driving force of water loss from a leaf-- tends to be greater in upper-canopy and canopy gaps than in the understory [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017; @niinemetsPhotosyntheticAcclimationSimultaneous2004b]. Increased tree height [upto 20m with higher LAI] can also lower mean and max temperature and $VPD$ in the understory, beyond a certain height the offset can plateau [@hardwickRelationshipLeafArea2015; @juckerCanopyStructureTopography2018].

@@

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994]. During $gs$, evapotranspiration facilitates heat dissipation. But when leaf water demand exceeds water transport, driven by irradiance, $T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$ [@fauset_differences_2018].

@@

High wind speed (Fig. `r fig_NEON_vertical`) decreases $T_{leaf}$ - $T_{air}$ due to large boundary layer conductance ($g_a$)  [@daudetWindSpeedLeaf1999a]. Therefore, under hot conditions, high wind speeds and high $g_s$ would enable cooling but at lower wind conditions, thickness of air layer at the leaf surface curbs rapid leaf-to-air heat transfer, limiting $g_a$ and increasing $T_{leaf}$ - $T_{air}$ [@martin_boundary_1999; @leighInfluenceLeafSize2017; @robertsStomatalBoundaryLayerConductances1990a; @songTreeSurfaceTemperature2020]

Leaf interiors are $RH$ saturated [@buckleyHumidityLeavesWhy2019], hence when atmospheric $RH$ is lower, leaf to air evapotranspiration increases, decreasing $T_{leaf}$-$T_{air}$. Whereas when atmospheric $RH$ increases, leaf evapotranspiration and heat-dissapation is stifled, resulting in  greater $T_{leaf}$-$T_{air}$ [@songTreeSurfaceTemperature2020; @tibbittsHumidityPlants1979b; @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts].

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** Small leaves heat and cool faster due to smaller $ga$ than larger leaves that take longer due to large $ga$. Therefore, $T_{leaf}$-$T_{air}$ is lower for small leaves than large leaves that can exprience greater $T_{leaf}$-$T_{air}$. With adequate water, species with smaller leaves and larger $gs$ will be most effective at cooling during high temperatures than large leaf species [@songTreeSurfaceTemperature2020; @dongBiophysicalHomoeostasisLeaf2017; @konradLeafTemperatureIts; @leighInfluenceLeafSize2017; @leuzingerTreeSpeciesDiversity2007].

@@

The distribution of biochemical traits enable photosynthesis and metabolic processes, and protect against heat and light damage in sun-leaves. $N_{area}$, that is typically greater in the direction of new leaf growth [@niinemetsPhotosynthesisResourceDistribution2007a], optimizes to prevent photoinhition in along sun-leaves and, shade-leaves during sunflecks [@kitaoCanopyNitrogenDistribution2018]. Similarly, carotenoids and xanthophyll pigments-violaxanthin, antheraxanthin and zeaxanthin [$VAZ$], an antioxidant scavenging function- convert excess light to heat dissaptation [@mathurPhotosyntheticEfficiencySun2018a; @niinemetsAcclimationHighIrradiance1998]. Constraints to biochemical traits, particularly $VAZ$ can curb increasing temperature acclimation [@kumarathungeAcclimationAdaptationComponents2019a]. Additionally, isoprenes in producing species facilitate higher heat tolerance in exposed-canopies, making these species particularly advantageous [*REF*].

@@

**We have very little evidence as to how the temperature sensitivity of photosynthesis compares between sun and shade leaves.**
Biophysically, we might expect sun-leaves to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*), where earlier evidence showed that $T_{opt}$ of $J_{max}$ increased with irradiance in an upper-canopy leaf compared to shaded lower-canopy leaf [@niinemetsPhotosyntheticAcclimationSimultaneous2004b]. However, such a trend is not apparent among recent studies, where for 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [Slot et al. 2019 ; @hernandezSimilarTemperatureDependence2020]. Similarly, $T_{opt}$ of $V_{max}$ and $J_{max}$ didnot vary for tropical sun- and shade-leaves  [@slotPhotosyntheticHeatTolerance2019a; @hernandezSimilarTemperatureDependence2020], along the vertical gradient of tropical and temperate trees [@mau_temperate_2018; @millerOnlySunlitLeaves2021;@mau_temperate_2018].

Interestingly, within canopy Japanese oak sun and shade leaves under respective peak light conditions were found to maintain similar rate of excess energy production. Excessive light energy not utilized can cause photoinhibition.Thus, sun leaves utilize this excess energy through high electron transport rate ($ETR$) and thermal dissapation, whereas shade leaves compensate for low $ETR$ by adjusting their $N_{area}$ to maintain high photosythetic capacity than those required to maximize C net gain, a suggested mechanism for cirucmventing photoinihition and thermal dissaptation under sunflecks [@kitaoLeavesJapaneseOak2012; @kitaoCanopyNitrogenDistribution2018]. Perhaps, differences in these thermal stablization strategies realizes similar $T_{opt}$ for photosynthesis in sun and shade leaves.
